---
title: "Exploratory Data Analysis Using the ToothGrowth Dataset"
output: 
  html_document:
    keep_md: true
---

## Loading ToothGrowth Data

```{r loadData, echo=TRUE}
library(dplyr)
library(ggplot2)
library(datasets)
data(ToothGrowth)
```

## Basic Exploratory Data Analyses

```{r dataSynopsis, echo=TRUE}
str(ToothGrowth)
head(ToothGrowth)
tail(ToothGrowth)
```

### Number of Rows and Columns
```{r numRowCol, echo=TRUE}
dim(ToothGrowth)
```

### Sample Size (n)
```{r sampSize, echo=TRUE}
length(ToothGrowth$len)
```

### Mean Group by Dose, OJ and VC
```{r meanGroup, echo=TRUE}
aggregate(ToothGrowth$len,list(ToothGrowth$supp,ToothGrowth$dose),mean)
```

### Standard Deviation Group by Dose, OJ and VC
```{r stdGroup, echo=TRUE}
aggregate(ToothGrowth$len,list(ToothGrowth$supp,ToothGrowth$dose),sd)
```

### Boxplot Graph of Tooth Length versus Dose
```{r plot1, echo=TRUE, fig.align='center'}
ggplot(ToothGrowth, aes(x = factor(dose), y = len, fill = factor(dose))) + 
        geom_boxplot() + facet_grid(.~supp) + 
        labs(title = "Tooth Length vs. Dose  by for OJ & VC", 
             x = "Doses", y = "Tooth Length")

```

## Data Summary

```{r dataSummary, echo=TRUE}
summary(ToothGrowth)
```

```{r dataTable, echo=TRUE}
table(ToothGrowth$supp,ToothGrowth$dose)
```

## Comparison of Tooth Growth by Supp and Dose

Based on the box plot generated earlier, OJ appears to be doing better with 
dose 0.5 and 1 on tooth growth than VC. To test this hypothesis by hold 
a the mean of OJ and VC does not cross zero.

### Dose 0.5:
We are 95% confident that the limits of 1.719057 and 8.780943 actually do 
contain the difference between the two population means. Because those limts do 
not contain zero, this confidence interval suggests that it is very possible 
that the two population means are not equal.

```{r dose0.5, echo=TRUE}
ojdose0.5 <- ToothGrowth %>% filter(supp=="OJ" & dose=="0.5")
vcdose0.5 <- ToothGrowth %>% filter(supp=="VC" & dose=="0.5")
t.test(ojdose0.5$len,vcdose0.5$len)
```

### Dose 1:
We are 95% confident that the limits of 2.802148 and 9.057852 actually do 
contain the difference between the two population means. Because those limts do 
not contain zero, this confidence interval suggests that it is very possible 
that the two population means are not equal.

```{r dose1, echo=TRUE}
ojdose1 <- ToothGrowth %>% filter(supp=="OJ" & dose=="1")
vcdose1 <- ToothGrowth %>% filter(supp=="VC" & dose=="1")
t.test(ojdose1$len,vcdose1$len)
```

### Dose 2:
We are 95% confident that the limits of -3.79807 and 3.63807 actually do 
contain the difference between the two population means. However, because those 
limts do contain zero, this confidence interval suggests that it is very 
possible that the two population means are equal.

```{r dose2, echo=TRUE}
ojdose2 <- ToothGrowth %>% filter(supp=="OJ" & dose=="2")
vcdose2 <- ToothGrowth %>% filter(supp=="VC" & dose=="2")
t.test(ojdose2$len,vcdose2$len)
```

## Conclusion

We are 95% confident that dose 0.5 and dose 1 of OJ result in longer tooth 
length than dose 0.5 and dose 1 of VC. However, at the highest dose of 2, there 
is no statistically significant difference between the effects of OJ and VC.